---
title: "data-viz"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(leaflet)
library(dplyr)
library(tidyr)
library(ggplot2)
library(DT)
library(scales) # install.packages("scales")
library(patchwork) # install.packages("patchwork")
```

# Loading data

```{r}
data_url <- "https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Af119a05b-bbe7-4aea-93c6-85434dcb1c5e"

esc <- tryCatch(
    read.csv("data/escapement.csv", stringsAsFactors = FALSE),
    error=function(cond) {
        message(paste("Escapement file does not seem to exist, so get it from the KNB."))
        esc <- read.csv(url(data_url, method = "libcurl"), stringsAsFactors = FALSE)
        return(esc)
    }
)

head(esc)
```

# Data processing
```{r}
annual_esc <- esc %>%
  separate(sampleDate, c("Year", "Month", "Date"), sep= "-") %>%
  mutate(Year = as.numeric(Year), 
         Month = as.numeric(Month)) %>%
  group_by(Species, SASAP.Region, Year) %>%
  summarise(escapement = sum(DailyCount)) %>%
  filter(Species %in% c("Chinook", "Chum", "Coho", "Pink", "Sockeye"))

head(annual_esc)
  
```

# Data viz

## First look at the data
```{r}
ggplot(annual_esc, aes(x = Species, y = escapement)) +
  geom_col()
```

Changing color of bars

Everything that goes into aes is the variable mapping
```{r}
ggplot(annual_esc, aes(x = Species, y = escapement)) +
  geom_col(fill = "blue")
```

Per region:
```{r}
ggplot(annual_esc, aes(x = Species, y = escapement, fill = SASAP.Region)) +
  geom_col()
```

### Looking at the Kodiak region

```{r}
## need to use color instead of fill for a geom_line() if you want to add a third dimension to a geom_line()
annual_esc %>%
  filter(SASAP.Region == "Kodiak") %>%
  ggplot(aes(x = Year, y = escapement/10^6, color = Species)) + 
  geom_line() +
  geom_point() +
  ggtitle("Salmon annual escapement") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5), legend.position=c(0.3, 0.75), legend.title = element_blank())
  #facet_wrap(~Species)
```

### Looking at the Kodiak region

```{r}
## need to use color instead of fill for a geom_line() if you want to add a third dimension to a geom_line()
annual_esc %>%
  ggplot(aes(x = Year, y = escapement/10^6, color = Species)) + 
  geom_line() +
  geom_point() +
  ggtitle("Salmon annual escapement") +
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5), legend.position="bottom", legend.title = element_blank())+
  facet_wrap(~SASAP.Region, scales = "free_y")
```

## Use patchwork library to arrange multiple subplots
```{r}
p1 <- ggplot(annual_esc, aes(x = Species, y = escapement)) +
  geom_col()
p2 <- ggplot(annual_esc, aes(x = Species, y = escapement)) +
  geom_col(fill = "blue")

p1 + p2 + plot_layout(nrow = 2)

```

```{r}
percent <- function(percent_in, total){
  out <- percent_in*total/100
  #paste(as.character(out),, sep="")
  out
}

percent(10, 100)
```

Make a ratio factor an argument so that we can also use it to compute 1/1000; make it compute percentage by default
```{r}
#' Compute a ratio
#'
#' @param percent_in number used in the numerator of the ratio
#' @param total number value used to apply ratio
#' @param place optional - number used in the denominator of the ratio
#'
#' @return 
#' @export
#'
#' @examples ratio(10,55)

ratio <- function(percent_in, total, place=100){
  out <- percent_in*total/place
  out
}

ratio(10, 100, 1000)
```

To source your function from another Rscript file:

Save your function as an R script
Use source() function. This runs the script and loads everything in your global environment
```{r}
source("myTestFunction.R")
```

