---
title: "Geospatial analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(sf)
library(dplyr)
library(ggplot2)
library(scales)
library(ggmap)
```

# Read in data
```{r}
ak_regions <- read_sf("data/ak_regions_simp.shp")

class(ak_regions)
plot(ak_regions)
head(ak_regions)
```

transform data
```{r}
# name different transformations of your object using the EPSG code
ak_regions_3338 <- ak_regions %>%
  st_transform(crs=3338)

plot(ak_regions_3338)
```

Filter data using tidyverse functions
```{r}
ak_regions_3338 %>%
  ## choose only rows that equal Southeast region
  filter(region == "Southeast") %>%
  ## pick out just the region column
  select(region)

## the geometry is retained
```

# Spatial joins and summaries
```{r}
## read in population csv that has city name and lat long, but we want to know which region each city falls within
pop <- read.csv("data/alaska_population.csv", stringsAsFactors = FALSE)
head(pop)

class(pop)
```

Need to convert data frame to sf object and has to be in the same projection in order to do a spatial join
```{r}
# Assume that it is WGS84
# use st_as_sf to force it to be a sf object
pop_4326 <- st_as_sf(pop, 
                     coords = c("lng", "lat"),
                     crs = 4326,
                     ## keep or remove the lat/long coordinates
                     remove = FALSE
                     )

head(pop_4326)

## transfrom 4326 to 3338 in order to perform join
pop_3338 <- pop_4326 %>%
  st_transform(crs = 3338)
```

```{r}
pop_joined <- st_join(pop_3338, ak_regions_3338, join = st_within)
head(pop_joined)
plot(pop_joined)
```

## Calculate total population by region
```{r}
pop_region <- pop_joined %>%
  as.data.frame() %>%
  group_by(region) %>%
  summarize(total_pop = sum(population))

head(pop_region)
```


```{r}
pop_region_3338 <- left_join(ak_regions_3338, pop_region, by="region")
head(pop_region_3338)
plot(pop_region_3338)
```

Create maps for mgmt region
```{r}
pop_mgmt_3338 <- pop_region_3338 %>%
  group_by(mgmt_area) %>%
  summarize(total_pop = sum(total_pop), do_union=FALSE)

head(pop_mgmt_3338)
plot(pop_mgmt_3338["total_pop"])
```

```{r}
## write out 
write_sf(pop_region_3338, "data/ak_regions_pop.shp", delete_layer = TRUE)
```


# Make maps
```{r}
rivers_3338 <- read_sf("data/ak_rivers_simp.shp")
st_crs(rivers_3338)
```

```{r}
## ggnewscale package allows you set scales for multiple geom_sf calls
## use get_legend from cowplot
## insert <ggnewscale::newscale("size")> between geom_sf() and scale_size()

ggplot()+
  geom_sf(data = pop_region_3338, aes(fill = total_pop)) +
  geom_sf(data = rivers_3338, aes(size = StrOrder), color = "black") +
  ## scale_size applies only to rivers because it's the only one that has the size argument in the aes()
  scale_size(range = c(0.01, 0.2), guide = F) + 
  geom_sf(data = pop_3338, aes(), size = 0.5) +
  #geom_sf(data = pop_3338, aes(size = population), size = 0.5) +
  #scale_size(range = c(0.1, 0.5), guide = F) + 
  theme_bw() +
  labs(fill = "Total population")+
  ## comma is a function that comes from scale package
  scale_fill_continuous(low = "khaki", high = "firebrick", labels = comma)
```


# Getting basemaps using ggmap
```{r}
## first transform to the pseudo mercator projection which is used by google maps, open street maps
pop_3857 <- pop_3338 %>%
  st_transform(crs = 3857)

pop_3857


## use stamen_map
```

# function to fix the ggmap bug:
```{r}
# Define a function to fix the bbox to be in EPSG:3857
# See https://github.com/dkahle/ggmap/issues/160#issuecomment-397055208
ggmap_bbox_to_3857 <- function(map) {
  if (!inherits(map, "ggmap")) stop("map must be a ggmap object")
  # Extract the bounding box (in lat/lon) from the ggmap to a numeric vector, 
  # and set the names to what sf::st_bbox expects:
  map_bbox <- setNames(unlist(attr(map, "bb")), 
                       c("ymin", "xmin", "ymax", "xmax"))
  
  # Coonvert the bbox to an sf polygon, transform it to 3857, 
  # and convert back to a bbox (convoluted, but it works)
  bbox_3857 <- st_bbox(st_transform(st_as_sfc(st_bbox(map_bbox, crs = 4326)), 3857))
  
  # Overwrite the bbox of the ggmap object with the transformed coordinates 
  attr(map, "bb")$ll.lat <- bbox_3857["ymin"]
  attr(map, "bb")$ll.lon <- bbox_3857["xmin"]
  attr(map, "bb")$ur.lat <- bbox_3857["ymax"]
  attr(map, "bb")$ur.lon <- bbox_3857["xmax"]
  map
}
```

```{r}
bbox <- c(-170, 52, -130, 64) ## southeast Alaska
## bbox argument in get_stamenmap in lat/long
ak_map <- get_stamenmap(bbox, zoom = 4)
ak_map_3857 <- ggmap_bbox_to_3857(ak_map)

class(ak_map_3857)
```

```{r}
ggmap(ak_map_3857) + 
  ## ggmap function is setting many mapping aesthetics so you have to override by setting inherit.aes = FALSE
  geom_sf(data = pop_3857, aes(color = population), inherit.aes = F) +
  scale_color_continuous(low = "khaki", high = "firebrick", labels = comma)
```

# Create an interactive map

## First define leaflet projection
```{r}
epsg3338 <- leaflet::leafletCRS(
  crsClass = "L.Proj.CRS",
  code = "EPSG:3338",
  proj4def =  "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs",
  resolutions = 2^(16:7))
```

## reproject the pop region object to wgs84
```{r}
pop_region_4326 <- pop_region_3338 %>%
  st_transform(crs = 4326)
```

```{r}
pal <- colorNumeric(palette = "Reds", domain = pop_region_4326$total_pop)

m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
        addPolygons(data = pop_region_4326, 
                    fillColor = ~pal(total_pop),
                    weight = 1,
                    color = "black",
                    fillOpacity = 1,
                    label = ~region) %>% 
        addLegend(position = "bottomleft",
                  pal = pal,
                  values = range(pop_region_4326$total_pop),
                  title = "Total Population")

m
```

```{r}
pal <- colorNumeric(palette = "Reds", domain = pop_region_4326$total_pop)

m <- leaflet() %>%
        addTiles() %>%
        addPolygons(data = pop_region_4326, 
                    fillColor = ~pal(total_pop),
                    weight = 1,
                    color = "black",
                    fillOpacity = 1,
                    label = ~region) %>% 
        addLegend(position = "bottomleft",
                  pal = pal,
                  values = range(pop_region_4326$total_pop),
                  title = "Total Population")

m
```

# Raster analysis

Reading the data
```{r}
library(raster) ## required sp

lc_3338 <- raster("/home/shares/scientist/ds-workshop/ak_nlcd_2011_landcover_50m_3338.tif")

lc_3338
```

Plot
```{r}
plot(lc_3338)
plot(pop_3338, add = TRUE)
```

## extract most frequent land cover
```{r}
raster_points <- extract(lc_3338, pop_3338, buffer = 500, fun = modal)
raster_points
```

```{r}
pop_joined$land_cover <- raster_points

# Read the land cover legend
Legend_lc <- read.csv("/home/shares/scientist/ds-workshop/legend_ak_nlcd_2011.csv", stringsAsFactors = FALSE)
Legend_lc
```

```{r}
# It is a lot of categories, let us consolidate this
Legend_lc <- Legend_lc %>%
  mutate(main_lc = ifelse(ID %in% 40:49, "Forest",
                              ifelse(ID %in% 20:29, "Urban",
                                     ifelse(ID %in% 50:59, "Shrub",
                                             ifelse(ID %in% 70:79, "Grass",
                                                     ifelse(ID %in% 80:89, "Crops",
                                                             ifelse(ID %in% 90:99, "Wetland", Land.Cover.Type)
                                                             )
                                                     )
                                             )
                                     )
                              )
  )

Legend_lc

# Join the LC categories to the population data
pop_3338_cover <- left_join(pop_joined, Legend_lc, by=c("land_cover"="ID")) %>% 
  dplyr::select(-Red, -Green, -Blue, -Land.Cover.Type)

pop_3338_cover
```

## Plot
```{r}
# Create color palette by keeping last color of each group
pal <- Legend_lc %>% 
  group_by(main_lc) %>% 
  slice(n()) %>% # Keeping the last color of the groups
  ungroup %>% 
  arrange(ID) %>%
  mutate(color_hex = rgb(Red, Green, Blue, max = 255)) %>% 
  dplyr::select(main_lc, color_hex)

# turn pallete into a list for plotting
pal_list <- pal$color_hex
names(pal_list) <- pal$main_lc

pal_list
```

```{r}
ggplot(pop_3338_cover, aes(region, population, fill = main_lc)) +
    geom_col() +
    scale_y_continuous(labels = comma) +
    scale_fill_manual(values = pal_list) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    ggtitle("Land cover type by region and population") +
    labs(fill = "", y = "Population", x = "")
```

Create a map
```{r}
ggplot() +
    geom_sf(data = ak_regions_3338, aes(), color = "black") +
    geom_sf(data = pop_3338_cover, aes(color = main_lc,
                                       size = population), show.legend = "point") +
    scale_size_continuous(guide = F) +
    scale_color_manual(values = pal_list) +
    theme_bw() +
    theme(legend.position = "bottom", legend.title = element_blank())
```

## cropping the Copper River Region

```{r}
copper_region_3338 <- ak_regions_3338 %>%
  filter(region == "Copper River") 

plot(copper_region_3338)
```

Crop raster
```{r}
copper_lc_3338 <- crop(lc_3338, copper_region_3338)

plot(copper_lc_3338)
```

Defining forest mask function
```{r}
forest_masker <- function(x){
  x[x < 40 | x > 49] <- 0
  x[x >= 40 & x <= 49] <- 1
  x
}
```

```{r}
copper_forested_3338 <- calc(copper_lc_3338, forest_masker)

plot(copper_forested_3338)
```

Selecting the copper river population
```{r}
copper_pop_3338 <- pop_3338_cover %>%
  filter(region == "Copper River")

```

```{r}
forested_count <- extract(copper_forested_3338, copper_pop_3338, buffer = 500, fun = sum)
forested_count
```

Calculate percent forest cover 
20*20 is the number of pixels within the 500 m buffered area
100 is to express value as percentage
```{r}
copper_pop_3338$forest_cov <- 100*forested_count/(20*20)

copper_pop_3338$forest_cov
```

```{r}
ggplot() +
    geom_sf(data = copper_region_3338, aes(), color = "black") +
    geom_sf(data = copper_pop_3338, aes(color = forest_cov,
                                       size = population)) +
    scale_size_continuous(guide = F) +
    scale_colour_gradientn(colours = terrain.colors(10, rev=TRUE)) +
    theme_bw() +
    theme(legend.position = "bottom") +
    labs(color = "Percent Forest Cover")
```

