---
title: "DataCleaning"
output: html_document
---

```{r setup, include=FALSE}
## echo = false will not show code but will show results
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
```

# Read in data

We are using data from [Mike Byerly. Alaska commercial salmon catches by management region (1886- 1997). Gulf of Alaska Data Portal. df35b.304.2.](https://knb.ecoinformatics.org/view/df35b.304.2)

```{r}
catch_original <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", method = "libcurl"), stringsAsFactors = FALSE)

head(catch_original)
```


# Clean and reshape
%>% (ctrl shift m) = pipe argument takes the previous output as the first arguement in the next function
remove marginal sum and notes columns using 'select()'

```{r}
catch_clean <- catch_original %>% 
  #select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)
  select(-All, -notesRegCode) %>%
  ## clean up the data after QA to convert I to 1
  mutate(Chinook = ifelse(Chinook == "I", 1, Chinook)) %>%
  mutate(Chinook = as.numeric(Chinook))

summary(catch_clean)
```

## QA 
Find where as.numeric couldn't convert Chinook to a number

```{r}
## where 
i <- which(is.na(catch_clean$Chinook))
i
```
```{r}
catch_original[i,]
```
original data for that row should be a 1 not an I

# Reshape from wide to tall format using 'pivot_longer' (from dplyr)
Similar to melt and cast (in plyr) and gather and spread from and older version of dplyr.
It is better because it can do splitting across multiple columns unlike melt/cast

Mutate = create new column and if you give it the same name it will overwrite the original column values. 

rename = just renames column
```{r}
catch_long <- catch_clean %>%
  pivot_longer(col = -c(Region, Year), 
               names_to = "species",
               values_to = "catch") %>%
  # mutate(catch = catch * 1000) %>%
  rename(catch_thousands = catch) %>%
  mutate(catch = catch_thousands *1000) %>%
  select(-catch_thousands)

head(catch_long)

```

example with pivot_wider
```{r}
catch_wide <- catch_long %>% 
  pivot_wider(names_from = Year, 
              values_from = catch)

head(catch_wide)
```

## group_by and summarize
```{r}
mean_region <- catch_long %>%
  group_by(Region, species) %>%
  summarise(catch_mean = mean(catch), 
            num_obs =n())

head(mean_region)
```

```{r}
mean_sp <- catch_long %>%
  group_by(species) %>%
  summarise(catch_mean = mean(catch), 
            num_obs = n())

head(mean_sp)
```

In what year was the total salmon catch the highest?

```{r}
sum_year <- catch_long %>%
  group_by(Year) %>%
  summarise(catch_total = sum(catch), 
            num_obs =n()) %>%
  arrange(-catch_total)

head(sum_year)
sum_year[which(sum_year$catch_total == max(sum_year$catch_total)),]
```

What year had the highest sockeye catch?
```{r}
annual_sockeye <- catch_long %>%
  filter(species == "Sockeye") %>%
  group_by(Year) %>%
  summarise(catch_total = sum(catch), 
            num_obs =n()) %>%
  arrange(-catch_total)

head(annual_sockeye)
max <- max(annual_sockeye$catch_total)
max
annual_sockeye[which(annual_sockeye$catch_total == max),]
```



# Join to regions table

Read in regions data
```{r}
region_defs <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1", stringsAsFactors = FALSE) %>%
  select(code, mgmtArea)

head(region_defs)
```

```{r}
catch_joined <- left_join(catch_long, region_defs, by = c("Region" = "code"))
head(catch_joined)
```

